---
title: "W2 Lab: Worked Answers"
subtitle: "Text as Data | National Taiwan University"
format:
  html:
    embed-resources: true
    toc: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

## Before starting

These answers include all worked examples and solutions to the **11 practice exercises**. Try [the student handout](../Lab_Session_W2.html) first. Run setup below once, then continue in order. The supplied data and packages are the same as in the student version.

## Setup: packages and data

```{r load_packages}
# First use only: uncomment these lines to install the packages.
# install.packages(c("dplyr", "stringr", "ggplot2", "quanteda",
#                    "quanteda.textstats", "quanteda.textplots"))
library(dplyr)
library(stringr)
library(ggplot2)
library(quanteda)
library(quanteda.textstats)
library(quanteda.textplots)
```

```{r read_hoc_sample}
# Find the supplied data. If asked, choose the .rds file in W2/lab/Data.
paths <- c("W2/lab/Data/hc_sample_1945_2025.rds",
           "lab/Data/hc_sample_1945_2025.rds",
           "../Data/hc_sample_1945_2025.rds",
           "Data/hc_sample_1945_2025.rds")
found <- paths[file.exists(paths)]
if (length(found) == 0) found <- file.choose()
data_path <- normalizePath(found[1])
setwd(dirname(dirname(data_path)))
hoc <- readRDS(data_path)
nrow(hoc)
```

**Read the result:** `5000` is the number of speech contributions, not the number of MPs. The same MP can contribute more than once. Keep this R session open for the rest of the lab.

Create the `year` variable from the speech date.

```{r create_year}
hoc <- hoc |>
  mutate(year = as.integer(format(date, "%Y")))

hoc |>
  select(date, year, speaker, party, agenda, terms) |>
  head(5)
```

## Reading the R code

Select a complete block and use `Ctrl+Enter` (Windows/Linux) or `Cmd+Enter` (Mac) in RStudio. Run the blocks in order, inspect the output, then try changing one input.

| Syntax | Meaning | Example |
|---|---|---|
| `<-` | Save a result under a name | `hoc_corpus <- corpus(hoc, text_field = "text")` |
| `$` | Select a column | `hoc$party` |
| `[1:8]` | Select the first eight items | `hoc$agenda[1:8]` |
| <code>&#124;&gt;</code> | Pass the result to the next function | <code>hoc &#124;&gt; head(5)</code> |
| `==`, `>=`, `&` | Equal to, at least, and both conditions | `party == "Labour" & year >= 2010` |
| `head(x, 5)` | Show a few results | `head(agenda_labels, 5)` |

For a pipe, read each line as “then”: `hoc |> select(party) |> head(5)` means take `hoc`, select the party column, then show five rows. The examples use ready-made package functions; you do not need to write functions or loops.

### Functions you will reuse

| Task | Function |
|---|---|
| Search, extract or replace text | `str_detect()`, `str_extract()`, `str_replace_all()` |
| Keep rows or count categories | `filter()`, `count()` |
| Create and subset a corpus | `corpus()`, `corpus_subset()` |
| Change the document unit | `corpus_reshape()` |
| Split text and combine phrases | `tokens()`, `tokens_compound()` |
| Inspect surrounding words | `kwic()` |
| Count features and combine documents | `dfm()`, `dfm_group()` |
| Inspect frequency and group differences | `topfeatures()`, `textstat_keyness()` |

## 1. Inspect and clean text

R stores text as character data. The House of Commons sample contains several character variables, including `speaker`, `party`, `agenda`, `text`, and `speech_url`.

Let us start with a small vector of agenda labels. These are useful for learning string operations because parliamentary agenda labels are often messy: they vary in capitalisation, contain repeated phrases, and sometimes include bracketed contextual information.

```{r create_agenda_vector}
agenda_labels <- hoc$agenda[1:8]

agenda_labels
str(agenda_labels)
```

The `stringr` package contains many useful functions for working with character data. We will start with basic pattern matching.

### Extracting and locating patterns

`str_extract()` returns the first match to a pattern.

```{r extract_first_pattern}
str_extract(agenda_labels, "Trade")
str_extract(agenda_labels, "WAR")
```

Pattern matching is case-sensitive by default. Compare the following:

```{r case_sensitive_matching}
str_extract(agenda_labels, "war")
str_extract(str_to_lower(agenda_labels), "war")
```

**Read the result:** `NA` means the pattern was not found in that label. It does not mean the whole document is missing. Lowercasing makes the search insensitive to the original capitalization.

`str_which()` returns the positions of strings that contain a match.

```{r locate_matching_strings}
str_which(str_to_lower(agenda_labels), "war")
str_which(str_to_lower(agenda_labels), "trade")
```

`str_locate()` gives the start and end position of the first match inside each string.

```{r locate_pattern_positions}
str_locate(str_to_lower(agenda_labels), "war")
```

`str_locate_all()` gives the positions of all matches.

```{r locate_all_pattern_positions}
str_locate_all(str_to_lower(agenda_labels), "war")
```

### Regular expressions

Regular expressions are a compact language for describing text patterns. They are extremely useful for cleaning and searching text.

The expression `\\d` matches a digit. The expression `\\d+` matches one or more digits.

We can see this with parliamentary URLs, which contain dates and identifiers.

```{r inspect_speech_urls}
speech_urls <- hoc$speech_url[1:5]

speech_urls
```

```{r extract_digits}
str_extract(speech_urls, "\\d")
str_extract(speech_urls, "\\d+")
str_extract_all(speech_urls, "\\d+")
```

The `+` means "one or more of the preceding pattern". So `\\d+` collects a whole run of digits rather than just the first digit.

We can also extract dates from the URLs. The pattern below searches for four digits, followed by a slash, three lower-case letters, another slash, and two digits.

```{r extract_url_dates}
str_extract(speech_urls, "\\d{4}/[a-z]{3}/\\d{2}")
```

Here are a few useful regex patterns:

- `[a-z]`: any lower-case letter;
- `[A-Z]`: any upper-case letter;
- `[A-Za-z]`: any upper-case or lower-case letter;
- `\\d`: any digit;
- `\\s`: any whitespace;
- `\\b`: a word boundary;
- `.`: any character;
- `\\.`: a literal full stop.

### Word boundaries

Word boundaries help us avoid accidental matches. For example, `war` also appears inside longer words such as `towards`. The pattern `\\bwar\\b` matches `war` as a separate word.

```{r word_boundary_examples}
example_text <- c(
  "The war changed British politics.",
  "The committee moved towards a compromise.",
  "Post-war trade was discussed."
)

str_detect(str_to_lower(example_text), "war")
str_detect(str_to_lower(example_text), "\\bwar\\b")
```

**Read the result:** the broad pattern matches all three strings. The word-boundary pattern excludes `towards` but still matches `post-war`. Choose the rule that fits the intended meaning.

Notice that `post-war` still contains `war` after a hyphen. Hyphens are treated as boundaries by this pattern. Whether that is desirable depends on the research question.

### Replacing patterns

`str_replace()` replaces the first match. `str_replace_all()` replaces every match.

```{r replace_patterns}
str_replace(agenda_labels, "Trade", "TRADE")
str_replace_all(agenda_labels, "[aeiou]", "-")
```

In real QTA work, replacement is often used to standardise metadata. For example, the sample contains several versions of "Business of the House".

```{r inspect_business_agendas}
hoc |>
  filter(str_detect(str_to_lower(agenda), "business of the house")) |>
  count(agenda, sort = TRUE) |>
  head(10)
```

We can create a cleaned agenda variable by:

- converting to lower case;
- removing extra whitespace;
- changing title case variants into a common label;
- removing bracketed contextual information when we want a shorter label.

```{r clean_agenda_labels}
hoc <- hoc |>
  mutate(
    agenda_clean = agenda |>
      str_to_lower() |>
      str_squish() |>
      str_replace_all("\\s*\\[.*\\]", ""),
    agenda_clean = case_when(
      agenda_clean == "business of the house" ~ "Business of the House",
      agenda_clean == "points of order" ~ "Points of Order",
      agenda_clean == "topical questions" ~ "Topical Questions",
      str_detect(agenda_clean, "^engagements") ~ "Engagements",
      TRUE ~ str_to_title(agenda_clean)
    )
  )

hoc |>
  count(agenda_clean, sort = TRUE) |>
  head(15)
```

**Check:** compare `agenda` with `agenda_clean`. Which differences were standardized, and which details were removed? The original column is kept so that you can review the changes.

This kind of cleaning is not neutral. For example, collapsing all labels that start with `Engagements` into one category may be helpful for descriptive plots, but it also removes detail. Cleaning choices should be documented.

## 2. Build and inspect a corpus

So far we have worked with character vectors and data frames. The `quanteda` package lets us create a corpus object. A corpus stores texts and document-level metadata together.

```{r create_hoc_corpus}
hoc_corpus <- corpus(
  hoc,
  text_field = "text"
)
```

We can inspect the corpus using `summary()`.

```{r summarise_hoc_corpus}
summary(hoc_corpus, n = 10)
```

The number of documents in the corpus is:

```{r count_documents}
ndoc(hoc_corpus)
```

**Read the result:** the corpus still has 5,000 documents. Creating a corpus changes how the data are stored; it does not select or classify speeches.

The document variables, or `docvars`, contain the metadata.

```{r inspect_docvars}
names(docvars(hoc_corpus))
head(docvars(hoc_corpus, "party"), 10)
head(docvars(hoc_corpus, "agenda_clean"), 10)
```

We can inspect the text of a document by converting it to a character vector.

```{r inspect_corpus_text}
as.character(hoc_corpus)[1]
```

### Select documents

We can subset a corpus using `corpus_subset()`. The following code keeps Labour speeches since 2010.

```{r subset_labour_since_2010}
labour_since_2010 <- corpus_subset(
  hoc_corpus,
  party == "Labour" & year >= 2010
)

ndoc(labour_since_2010)
summary(labour_since_2010, n = 5)
```

**Read the result:** the new corpus keeps only documents satisfying both conditions. `hoc_corpus` remains the full sample.

We can also subset by string patterns in document variables. For example, let us keep speeches with agenda labels related to Prime Minister's Questions.

```{r subset_engagements}
engagements_corpus <- corpus_subset(
  hoc_corpus,
  agenda_clean == "Engagements"
)

ndoc(engagements_corpus)
summary(engagements_corpus, n = 5)
```

### Change the document unit

Sometimes we want a different unit of analysis. `corpus_reshape()` can reshape documents into sentences or paragraphs. This is useful when speech-level documents are too long for a coding task.

```{r reshape_to_sentences}
engagements_sentences <- corpus_reshape(
  engagements_corpus,
  to = "sentences"
)

ndoc(engagements_corpus)
ndoc(engagements_sentences)
```

In the reshaped corpus, each sentence is now treated as a document. This is a major research design choice: changing the unit of analysis can change the conclusions of an analysis.

```{r inspect_sentence_corpus}
as.character(engagements_sentences)[1:5]
```

**Check:** read the sentence boundaries around parliamentary abbreviations such as `right hon.`. A full stop may belong to an abbreviation rather than the end of a sentence. Inspect the actual output before treating every detected sentence as a complete statement.

## 3. Inspect tokens and phrases

Tokenization breaks text into smaller units such as words. We will tokenize the House of Commons corpus and preserve a few multiword expressions.

```{r tokenize_hoc_corpus}
hoc_tokens <- tokens(
  hoc_corpus,
  remove_punct = TRUE,
  remove_symbols = TRUE,
  remove_numbers = FALSE
)
```

Some political concepts are multiword expressions. If we want `Prime Minister` or `European Union` to be treated as phrases, we can compound them.

```{r compound_multiword_expressions}
hoc_tokens <- tokens_compound(
  hoc_tokens,
  pattern = phrase(c(
    "Prime Minister",
    "House of Commons",
    "European Union",
    "National Health Service"
  ))
)
```

Now we can inspect keywords in context using `kwic()`.

```{r kwic_european_union}
kwic(
  hoc_tokens,
  pattern = phrase("European_Union"),
  window = 8
) |>
  head(10)
```

We can also inspect a single-word pattern.

```{r kwic_economy}
kwic(
  hoc_tokens,
  pattern = "economy",
  window = 8,
  valuetype = "fixed",
  case_insensitive = TRUE
) |>
  head(10)
```

**Check:** read the words before and after each match. A mention of `economy` alone does not tell you whether the speaker evaluates it positively or negatively.

## 4. Count and compare features

A document-feature matrix, or DFM, represents documents as rows and features as columns. In most introductory QTA workflows, these features are words or phrases. The cell values usually count how often each feature appears in each document.

The previous section created tokens for KWIC inspection. Now we will create a second tokens object for DFM construction. For this workflow we lower-case the text and remove English stopwords. We keep `padding = TRUE` at first because this helps when identifying collocations.

```{r tokenize_for_dfm}
hoc_tokens_dfm <- tokens(
  hoc_corpus,
  what = "word",
  remove_punct = TRUE,
  padding = TRUE,
  remove_symbols = TRUE,
  remove_numbers = FALSE,
  remove_url = TRUE,
  remove_separators = TRUE,
  split_hyphens = FALSE
) |>
  tokens_tolower() |>
  tokens_remove(stopwords("en"), padding = TRUE)
```

### Collocations

Collocations are words that occur together more often than we would expect by chance. In parliamentary text, examples might include `prime minister`, `European Union`, or `hon. member`.

```{r find_collocations}
hoc_collocations <- textstat_collocations(
  hoc_tokens_dfm,
  min_count = 10
) |>
  arrange(desc(lambda))

head(hoc_collocations, 20)
```

We can compound strong collocations so that they are treated as single features.

```{r compound_collocations_for_dfm}
hoc_collocation_phrases <- hoc_collocations |>
  filter(lambda > 10) |>
  pull(collocation) |>
  phrase()

hoc_tokens_dfm <- tokens_compound(
  hoc_tokens_dfm,
  pattern = hoc_collocation_phrases
)

hoc_tokens_dfm <- tokens_remove(hoc_tokens_dfm, "")
```

### Create a DFM

Let us create a DFM and group it by party. This means that each row of the grouped DFM is a party, and each column is a feature.

```{r create_party_dfm}
hoc_dfm_party <- dfm(hoc_tokens_dfm)

hoc_dfm_party <- dfm_group(
  hoc_dfm_party,
  groups = docvars(hoc_dfm_party, "party")
)

dim(hoc_dfm_party)
```

**Read the result:** `dim()` returns rows, then columns. After grouping, rows represent parties rather than individual speeches. Columns represent features; cells add up their occurrences.

The `topfeatures()` function shows the most common features.

```{r inspect_top_features}
topfeatures(hoc_dfm_party, 25)
```

Most words do not appear in most documents. For this reason, DFMs are usually sparse. `quanteda` stores DFMs efficiently, but trimming rare features still often makes analyses easier to inspect.

```{r trim_party_dfm}
hoc_dfm_party_trimmed <- dfm_trim(
  hoc_dfm_party,
  min_termfreq = 10
)

dim(hoc_dfm_party)
dim(hoc_dfm_party_trimmed)
```

**Check:** compare the two dimensions. Trimming removes features below the threshold; it does not turn the remaining counts into percentages.

Trimming is not neutral. Rare words may be noise, but they may also be substantively important. 

### Visualise features

Wordclouds are a quick way to inspect frequent features. They are not usually the best final visualisation for research, but they are useful for a first look.

```{r plot_wordcloud}
set.seed(719102)  # keep the layout reproducible
textplot_wordcloud(hoc_dfm_party_trimmed, max_words = 50)
```

A frequency plot is often easier to read.

```{r plot_feature_frequency}
hoc_feature_frequency <- textstat_frequency(
  hoc_dfm_party_trimmed,
  n = 25
)

hoc_feature_frequency <- hoc_feature_frequency |>
  mutate(feature = reorder(feature, frequency))

ggplot(
  hoc_feature_frequency,
  aes(x = feature, y = frequency)
) +
  geom_point() +
  coord_flip() +
  labs(
    x = NULL,
    y = "Frequency",
    title = "Most frequent features in the party-level DFM"
  ) +
  theme_minimal()
```

### Compare parties using keyness

Keyness compares feature usage in a target group with feature usage in a reference group. Here we inspect which features are relatively distinctive of Conservative speeches compared with the other parties in the sample.

```{r conservative_keyness}
conservative_keyness <- textstat_keyness(
  hoc_dfm_party_trimmed,
  target = docnames(hoc_dfm_party_trimmed) == "Conservative"
)

head(conservative_keyness, 20)

textplot_keyness(conservative_keyness, n = 20)
```

**Read the result:** positive chi-squared keyness values indicate features relatively overrepresented in Conservative speeches; negative values favor the reference group. This is a comparison of usage, not a sentiment score.

### Track a feature over time

We can also create a DFM grouped by year. Here we focus on Conservative speeches and count how often the feature `economy` appears in each year.

```{r conservative_economy_by_year}
hoc_dfm_year_conservative <- dfm(hoc_tokens_dfm)

hoc_dfm_year_conservative <- dfm_subset(
  hoc_dfm_year_conservative,
  party == "Conservative"
)

hoc_dfm_year_conservative <- dfm_group(
  hoc_dfm_year_conservative,
  groups = docvars(hoc_dfm_year_conservative, "year")
)

docvars(hoc_dfm_year_conservative, "economy") <- as.numeric(
  hoc_dfm_year_conservative[, "economy"]
)

economy_by_year <- docvars(hoc_dfm_year_conservative)

ggplot(economy_by_year, aes(x = as.integer(year), y = economy)) +
  geom_line() +
  geom_point(size = 1) +
  labs(
    x = NULL,
    y = "Frequency of 'economy'",
    title = "Conservative mentions of economy in the teaching sample"
  ) +
  theme_minimal()
```

**Read the plot:** these are raw counts in the supplied sample. A higher point can reflect more sampled speech as well as more use of `economy`. It is not automatically stronger concern or support.

## Practice exercises

Reuse the worked blocks above. Change one input at a time and explain what the output means. Complete exercise 4 before 5, and 9 before 10–11.

1. Create a character vector called `speaker_names` that contains the first 20 speaker names in `hoc`. Use `str_detect()` to identify which contain `Sir`.

```{r exercise_detect_sir}
speaker_names <- hoc$speaker[1:20]

speaker_names
str_detect(speaker_names, "Sir")
speaker_names[str_detect(speaker_names, "Sir")]
```

2. Use `str_extract_all()` to extract all four-digit years from the first 20 `speech_url` values.

```{r exercise_extract_years_from_urls}
speech_urls_20 <- hoc$speech_url[1:20]

str_extract_all(speech_urls_20, "\\b\\d{4}\\b")
```

3. Create a new variable called `mentions_prime_minister` that detects the phrase `Prime Minister` in the speech text. How many speeches mention this phrase?

```{r exercise_prime_minister_indicator}
hoc <- hoc |>
  mutate(
    mentions_prime_minister = str_detect(
      str_to_lower(text),
      "\\bprime minister\\b"
    )
  )

sum(hoc$mentions_prime_minister)
```

4. Use `corpus_subset()` to create a corpus of Conservative speeches from 2016 onwards. How many documents does this corpus contain?

```{r exercise_subset_conservative}
conservative_since_2016 <- corpus_subset(
  hoc_corpus,
  party == "Conservative" & year >= 2016
)

ndoc(conservative_since_2016)
summary(conservative_since_2016, n = 5)
```

5. Reshape the Conservative corpus from exercise 4 to sentences. How many sentence-level documents does it contain?

```{r exercise_reshape_conservative_sentences}
conservative_sentences <- corpus_reshape(
  conservative_since_2016,
  to = "sentences"
)

ndoc(conservative_sentences)
as.character(conservative_sentences)[1:5]
```

6. Tokenize the full House of Commons corpus and preserve the multiword expressions `Prime Minister`, `European Union`, and `House of Commons`. Use `kwic()` to inspect `Prime_Minister`.

```{r exercise_kwic_prime_minister}
hoc_tokens_exercise <- tokens(
  hoc_corpus,
  remove_punct = TRUE,
  remove_symbols = TRUE,
  remove_numbers = FALSE
)

hoc_tokens_exercise <- tokens_compound(
  hoc_tokens_exercise,
  pattern = phrase(c(
    "Prime Minister",
    "European Union",
    "House of Commons"
  ))
)

kwic(
  hoc_tokens_exercise,
  pattern = phrase("Prime_Minister"),
  window = 8
) |>
  head(10)
```

7. The sample contains both `agenda` and `agenda_clean`. Count the 10 most common values of each. What changed after cleaning?

```{r exercise_compare_agenda_cleaning}
hoc |>
  count(agenda, sort = TRUE) |>
  head(10)

hoc |>
  count(agenda_clean, sort = TRUE) |>
  head(10)

# Cleaning collapses capitalization differences such as "BUSINESS OF THE HOUSE",
# "Business of the House", and "Business Of The House" into one label. It also
# collapses agenda labels beginning with "Engagements" into a common
# "Engagements" category.
```

8. Display the strongest three-feature collocations in `hoc_tokens_dfm`, ranked by `lambda`. Some features already contain compounded words, so these are not necessarily three original words.

```{r exercise_three_word_collocations}
textstat_collocations(
  hoc_tokens_dfm,
  size = 3,
  min_count = 5
) |>
  arrange(desc(lambda)) |>
  head(20)
```

9. Create a DFM from `hoc_tokens_dfm` and group it by `party`. Call the result `hoc_dfm_exercise`. How many documents and features does it have?

```{r exercise_create_grouped_dfm}
hoc_dfm_exercise <- dfm(hoc_tokens_dfm)

hoc_dfm_exercise <- dfm_group(
  hoc_dfm_exercise,
  groups = docvars(hoc_dfm_exercise, "party")
)

dim(hoc_dfm_exercise)
topfeatures(hoc_dfm_exercise, 20)
```

10. Trim `hoc_dfm_exercise` so that it only contains features that appear at least 20 times. Inspect the new dimensions.

```{r exercise_trim_grouped_dfm}
hoc_dfm_exercise_trimmed <- dfm_trim(
  hoc_dfm_exercise,
  min_termfreq = 20
)

dim(hoc_dfm_exercise)
dim(hoc_dfm_exercise_trimmed)
```

11. Use `textstat_keyness()` to display the most distinctive features for Labour speeches compared with all other parties.

```{r exercise_labour_keyness}
labour_keyness <- textstat_keyness(
  hoc_dfm_exercise_trimmed,
  target = docnames(hoc_dfm_exercise_trimmed) == "Labour"
)

head(labour_keyness, 10)
textplot_keyness(labour_keyness, n = 10)
```

## Sources

Adapted from Martijn Schoonvelde (2026), *QTA Lab 02: String Operations and Inspecting the House of Commons Corpus*, with acknowledgment of earlier materials by Stefan Müller. The original analytical examples and 11 exercises are retained; setup and explanations are adapted for NTU.
