---
title: "W4 Take-home Lab: LIWC-Format Dictionaries and Validation"
subtitle: "Text as Data | National Taiwan University"
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Before starting

**Ungraded take-home practice; no submission.** In this lab, we'll read a dictionary, change one of its rules, and check which sentences it gets right or wrong. Allow **45–60 minutes** for sections 1–4, with extra time if you are new to R. The party comparison is optional; the W6 preview takes about 10 additional minutes.

We'll use the 5,000 parliamentary speech contributions from W2 to look at words in context. We'll also use short example sentences to check our dictionary matches by hand. **You do not need to run W2 first:** the setup below loads the data.

Extract the lab ZIP, then open **`Lab_Session_W4.R`** in RStudio and read this handout alongside it. Keep `workflow.R`, `Dictionaries/` and `Data/` together. Run the whole setup block below once; it finds the lab folder or asks you to choose its `workflow.R`. Keep the same R session open. Install `quanteda` once; no LIWC purchase, model download or API is needed. You do not need to render the QMD.

```{r}
#| label: setup
# install.packages("quanteda")  # run once before class
library(quanteda)
# Find the lab folder; if asked, select workflow.R inside W4/lab.
paths <- c("W4/lab/workflow.R", "lab/workflow.R",
           "../lab/workflow.R", "../workflow.R", "workflow.R")
found <- paths[file.exists(paths)]
if (length(found) == 0) found <- file.choose()
setwd(dirname(normalizePath(found[1])))
source("workflow.R")  # load the data and prepared examples
```

Try the exercises before opening [the worked answers](answers/Lab_Session_W4_Answers.html). Bring questions to [instructor](https://calendly.com/ycliao/text-as-data-instructor) or [TA](https://calendly.com/ycliao/text-as-data-ta) office hours.

## Reading the R code

We'll use **quanteda functions and simple R arithmetic**. Work through each section from top to bottom in the same R session, checking the output as you go. In RStudio, use `Ctrl+Enter` (Windows) or `Cmd+Enter` (Mac) to run a selected line.

| Syntax | Meaning | Example |
|---|---|---|
| `<-` | Save a result under a name | `counts <- dfm(toks)` |
| `$` | Select one column | `test$human` |
| `[rows, columns]` | Select rows or columns | `counts[, "positive"]` |
| `==`, `!=` | Equal to, different from | `test$v2 != test$human` |
| `sum(x)` | Add values; TRUE counts as 1 | `sum(predicted == 1)` |
| `subset(x, condition)` | Keep matching rows | `subset(synthetic, split == "development")` |
| `%in%` | Check membership in a list | `doc_id %in% discovery_ids` |
| `&` | Both conditions must hold | `predicted == 1 & human == 1` |
| `head(x, 6)` | Display six rows | `head(contexts, 6)` |

`workflow.R` loads our texts, dictionaries and prepared examples. Run it through the setup block above; you can leave its functions as they are. We'll work with the resulting objects in the steps below.

### Functions you will reuse

| Task | Ready-made function |
|---|---|
| Build a corpus from a data table | `corpus()` |
| Split text into tokens | `tokens()` |
| Define or import a dictionary | `dictionary()` |
| Match phrases / match word counts | `tokens_lookup()` / `dfm_lookup()` |
| Inspect context | `kwic()` |
| Count words | `dfm()` and `ntoken()` |
| Optional: combine speeches by party | `dfm_group()` |
| Change counts to presence / absence | `dfm_weight(..., scheme = "boolean")` |
| Display a DFM as a regular table | `convert(..., to = "data.frame")` |

The five-sentence score table is prepared by setup. You will calculate a simple score and the validation measures below.

## LIWC and our teaching dictionary

**LIWC means Linguistic Inquiry and Word Count.** It counts words in categories such as emotion. In this lab, quanteda reads a small, original teaching dictionary saved in LIWC's legacy `.dic` format. It is not the official LIWC dictionary, and our score is not LIWC's Tone measure.

You only need quanteda. See [LIWC's explanation](https://liwc.app/help/liwc) and [quanteda's dictionary importer](https://quanteda.io/reference/dictionary.html) for background.

## 1. Read a dictionary and inspect its matches

**10 minutes.** Open the supplied ten-word list and see which words it matches. Between the two `%` lines are the category names; below the second `%`, each word has a category number.

```{r}
readLines("Dictionaries/teaching_affect.dic")
toy_sentiment <- dictionary(file = "Dictionaries/teaching_affect.dic",
                            format = "LIWC")
toy_sentiment
```

Let's start with two short texts. After importing the dictionary above, run this block:

```{r}
texts <- c("The policy is good.", "The policy is not good.")
example_tokens <- tokens(texts, remove_punct = TRUE)
example_dfm <- dfm(example_tokens)
example_hits <- dfm_lookup(example_dfm, toy_sentiment)
counts <- convert(example_hits, to = "data.frame")
counts
```

Both rows have one positive match. `tokens()` splits text into tokens; `dfm()` counts words; `dfm_lookup()` combines word counts into dictionary categories. The rule does not interpret `not`.

```{r}
word_count <- ntoken(example_tokens)
positive <- counts$positive
negative <- counts$negative
net_score <- 100 * (positive - negative) / word_count
net_score
```

**Scoring means turning dictionary matches into a number for each text.** We subtract negative from positive matches, divide by all word tokens, then multiply by 100. We get 25 and 20: one positive match divided by four and five words. The second score is lower because `not` adds a word to the denominator. Our rule still counts `good` as positive.

Now inspect the supplied examples that also include an unmatched text, balanced counts and an empty text. The table uses three summaries, where P = positive matches, N = negative matches and L = all word tokens:

| Column | Calculation | What to read |
|---|---|---|
| `net_per100` | 100 × (P − N) / L | Positive minus negative matches per 100 words |
| `coverage_per100` | 100 × (P + N) / L | Percentage of words matched by this small dictionary |
| `balance` | (P − N) / (P + N) | Among matched words, −1 = all negative; +1 = all positive |

A zero denominator gives `NA`: the score is undefined. The categories in this teaching dictionary do not overlap, and every entry is one word.

```{r}
# The setup above created these five named examples.
check_texts
check_tokens <- tokens(check_texts, remove_punct = TRUE)
check_dfm <- dfm(check_tokens)
check_hits <- dfm_lookup(check_dfm, toy_sentiment)
convert(check_hits, to = "data.frame")

# Prepared by workflow.R using the three formulas in the table above.
check_scores
```

**Check:** compare `good`, `negated`, `unmatched`, `balanced` and `empty`. Which examples have the same net score for different reasons? This table uses the supplied dictionary; it will not update automatically if you change the dictionary later.

**Discuss:** What would our rule need to do differently to distinguish "good" from "not good"? Keeping `not` in the tokens is only a first step; we still need a rule that uses it.

## 2. Build an energy-policy dictionary

**15–20 minutes.** Now we'll ask a different question: **does a speech discuss energy policy?** A relevant speech can be positive, negative or neither. We'll assign a dictionary prediction of 1 when at least one energy expression matches. To check these predictions, we'll read the texts ourselves using the rules below.


### How to label a speech {#energy-label-rules}

Read the whole speech contribution. Judge what it discusses, not whether a dictionary word appears.

- **1 — Relevant:** it discusses energy production, supply, prices, affordability or transition, such as electricity bills, heating costs, renewable energy or coal-mine closures. A substantive passage is enough; the speaker does not have to support the policy.
- **0 — Not relevant:** it does not discuss those issues. Political `power`, an unrelated legislative `bill`, personal `energy` and tear `gas` do not count. General climate discussion counts only when connected to energy.
- **Uncertain:** leave the label blank, write down why, and discuss it before scoring. Do not treat uncertainty as 0.

For each decision, record a short reason or supporting passage. Apply the same rules to texts with and without dictionary matches.

### Try the dictionary

```{r}
policy_entries <- read.csv("Dictionaries/energy_policy.csv")
head(policy_entries)
# Read the v1 patterns, then put them in a dictionary category.
v1_entries <- subset(policy_entries, version == "v1")
energy_v1 <- dictionary(list(energy = v1_entries$pattern))
v2_entries <- subset(policy_entries, version == "v2")
energy_v2 <- dictionary(list(energy = v2_entries$pattern))
energy_v1
energy_v2
contexts <- kwic(discovery_tokens, pattern = c("power", "gas", "bill*"),
                 valuetype = "glob", window = 8)
head(contexts, 10)
```

`Dictionaries/energy_policy.csv` stores version, category and pattern. V1 includes broad `power`, `gas` and `bill*`; illustrative v2 replaces several broad patterns with terms/phrases. `*` is a glob wildcard, not a regular expression. This CSV is a quanteda input, not a LIWC import file.

`discovery_tokens` excludes reserved real-text test IDs. It preserves punctuation and stopwords so phrases do not join across punctuation. A separate punctuation-free stream supplies word-token denominators. A phrase match counts as one expression, not all its constituent words. Read the original speech when KWIC is ambiguous.

To read the first KWIC result in full:

```{r}
first_id <- contexts$docname[1]
original <- subset(speeches, doc_id == first_id)
original$text
```

### Change one rule, then rerun

V1 includes `power`, which can mean political authority. Replace it with `wind farm*`, then compare the old and revised dictionaries on two teaching sentences:

```{r}
energy_revised <- dictionary(list(
  energy = c("energy", "wind farm*", "electric*", "gas", "bill*")
))
revision_texts <- c(
  wind = "The government should approve new wind farms.",
  authority = "Parliament must retain the power to question ministers."
)
revision_tokens <- tokens(revision_texts, remove_punct = FALSE)

before_tokens <- tokens_lookup(revision_tokens, energy_v1, capkeys = FALSE)
before_counts <- convert(dfm(before_tokens), to = "data.frame")
after_tokens <- tokens_lookup(revision_tokens, energy_revised, capkeys = FALSE)
after_counts <- convert(dfm(after_tokens), to = "data.frame")

data.frame(sentence = before_counts$doc_id,
           before = before_counts$energy,
           after = after_counts$energy)
```

**Read the result:** `wind` changes from 0 to 1; `authority` changes from 1 to 0. Our change fixes both examples. Next, we'll check other sentences to see which mistakes remain.

**Try:** in your own `energy_revised`, add or remove one entry, or replace a broad word with a phrase. Rerun the matching code after each change. Use `kwic(discovery_tokens, phrase("wind farm*"))` as a template to inspect your term in real context; read the original speech before deciding. Write down the term you changed, a supporting passage and one possible mistake.

For sections 3–4, we'll use the supplied **`energy_v2`** so we can compare our results with the worked answers. Your own changes remain in `energy_revised`; the scores we calculate next will refer to `energy_v2`.

To count multiword expressions, match the dictionary on the tokens first:

```{r}
matched_tokens <- tokens_lookup(discovery_tokens, energy_v2, capkeys = FALSE)
match_dfm <- dfm(matched_tokens)
head(match_dfm, 6)
```

Each value counts matched expressions in one speech. The original tokens remain available for context checks. Next, compare dictionary predictions with the labels on our teaching sentences.

## 3. Diagnose development errors

**10 minutes.** Let's look for mistakes before calculating a score. `Data/synthetic_validation.csv` contains 24 sentences written for this exercise, with labels in the column `human`. These are separate from the 5,000 parliamentary speeches. We'll inspect the 12 **development** sentences now and save the 12 **test** sentences for section 4.

```{r}
# Read the labelled teaching examples and generate predictions explicitly.
synthetic <- read.csv("Data/synthetic_validation.csv")
example_corpus <- corpus(synthetic, text_field = "text", docid_field = "doc_id")
example_tokens <- tokens(example_corpus, remove_punct = FALSE)

# Apply v1, then v2. Each matched phrase counts once.
v1_matches <- tokens_lookup(example_tokens, energy_v1, capkeys = FALSE)
v1_dfm <- dfm(v1_matches)
v1_binary <- dfm_weight(v1_dfm, scheme = "boolean")
v1_predictions <- convert(v1_binary, to = "data.frame")
synthetic$v1 <- v1_predictions$energy

v2_matches <- tokens_lookup(example_tokens, energy_v2, capkeys = FALSE)
v2_dfm <- dfm(v2_matches)
v2_binary <- dfm_weight(v2_dfm, scheme = "boolean")
v2_predictions <- convert(v2_binary, to = "data.frame")
synthetic$v2 <- v2_predictions$energy
```

`dfm_weight(..., scheme = "boolean")` turns a nonzero count into 1 and keeps zero as 0. `convert(..., to = "data.frame")` gives a regular table that is easy to inspect. These predictions come from our dictionary, while `human` contains the supplied teaching labels. Read development cases first; leave test cases for section 4.

```{r}
dev <- subset(synthetic, split == "development")
dev[, c("doc_id", "text", "human", "v1", "v2")]
```

**Try:** find one false positive and one false negative under v1. Which word caused the false match? Which expression did the dictionary miss? Check whether v2 fixes these cases, then find a mistake that remains.

```{r}
# Your error analysis here.
```

**Freeze means stop changing the rules before checking test results.** We'll keep the dictionary, labeling instructions, tokenization and prediction rule fixed (here, at least one match means 1). For your own project, save this version so you have a record of exactly what you tested.

## 4. Check a frozen rule on test examples

**10–15 minutes.** Finish the development discussion first. Use the supplied v2 as a fixed worked example. Predict which errors will lower precision or recall, then run the code.

```{r}
test <- subset(synthetic, split == "test")
# 0 = not relevant; 1 = relevant. Keep both categories in the table.
predicted <- factor(test$v2, levels = c(0, 1))
human <- factor(test$human, levels = c(0, 1))
table(predicted, human)
```

- **Precision = TP / (TP + FP):** how many predicted positives are correct?
- **Recall = TP / (TP + FN):** how many actual positives were found?
- **F1 = 2TP / (2TP + FP + FN):** how well are precision and recall balanced?

```{r}
# Compare each dictionary prediction with its teaching label.
TP <- sum(test$v2 == 1 & test$human == 1)
FP <- sum(test$v2 == 1 & test$human == 0)
FN <- sum(test$v2 == 0 & test$human == 1)
TN <- sum(test$v2 == 0 & test$human == 0)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
F1 <- 2 * TP / (2 * TP + FP + FN)
c(TP = TP, FP = FP, FN = FN, TN = TN)
c(precision = precision, recall = recall, F1 = F1)
```

**Read the result:** our dictionary flags 5 sentences, and 4 are relevant (precision = 4/5). Of the 6 relevant sentences, we find 4 (recall = 4/6). F1 combines precision and recall; it is not accuracy. If a denominator is zero in your own data, report that measure as `NA`.

These results describe the **12 test sentences**. To evaluate our dictionary on parliamentary speeches, we would need human labels for those speeches.

**Discuss:** How would we find missed cases if we inspected only dictionary matches? Why might accuracy look high when very few texts discuss energy? If we change v2 after reading the test errors, what would we need for a fresh evaluation? We'll return to these questions in W6.

## Before W6: five texts and one difficult decision

**10 minutes; no upload.** W5 is a holiday. Before we meet in W6, read these five parliamentary speeches from our development sample:

```{r}
development_texts <- subset(coding_template, split == "development")
preview <- head(development_texts, 5)
preview[, c("doc_id", "text")]
```

Read without dictionary predictions. Apply the [three labeling rules above](#energy-label-rules), record a short reason for each 0/1 label, and flag uncertainty instead of guessing. Bring one difficult decision, or a hypothetical boundary case if all five are clear. A random batch need not contain any positives.

| W4 preparation | W6 develops |
|---|---|
| Define a category and defend a label | Annotation quality and inter-coder reliability |
| Notice false positives and negatives | Performance evaluation and error analysis |
| Separate development from evaluation | Training/test splits and cross-validation |
| Keep a transparent dictionary baseline | Naive Bayes and SVM document classifiers |

We'll use these five speeches to discuss how we decide on a label. Read them without looking at dictionary predictions, and keep them in the development set if we use them to revise our rules. We'll need more labeled texts before training or evaluating a classifier.

## Real-text validation in W6

In W6, we'll work through human coding and validation with real texts. For now, complete the short-sentence exercise and the five-speech preview above. Setup also reserves **50 parliamentary speeches** for later testing. We leave those untouched in W4; they are a separate set from the 12 test sentences we just used.

## Optional: compare parties

This extension is not needed for the four core tasks.

<details>
<summary>Show the party comparison</summary>

After sections 1–4, you can summarize dictionary matches by party for the full sample. `policy_tokens` preserves punctuation for phrase matching; `toks` contains the original word tokens for the denominator:

```{r}
# quanteda adds counts within each party.
party_matches <- tokens_lookup(policy_tokens, energy_v2, capkeys = FALSE)
party_dfm <- dfm(party_matches)
party_dfm <- dfm_group(party_dfm, groups = party)

# Group original word counts separately to preserve the denominator.
party_words <- dfm(toks)
party_words <- dfm_group(party_words, groups = party)

party_summary <- convert(party_dfm, to = "data.frame")
party_summary$tokens <- ntoken(party_words)
party_summary$per1000 <- 1000 * party_summary$energy / party_summary$tokens
head(party_summary, 6)
```

`dfm_group()` performs the grouping; `convert()` turns the result into a table. We keep the final division explicit to show its denominator. Do not apply `dfm_weight(..., "prop")` to a matched-only DFM here: it would divide by matched counts instead of all original words.

For each party, we calculate **1,000 * total matches / total original word tokens**. We add the counts before dividing, rather than averaging the rates of individual speeches. These rates tell us how often our dictionary matches words in each party's sampled speeches. To interpret a difference, we also need to consider when the speeches were made, their topics, and how well the dictionary works. A match alone does not tell us whether the speaker supports an energy policy.

</details>

POS and TextRank examples are in the separate classroom materials. They are not part of this take-home lab.

## Reading connections

Our three required readings are Denny and Spirling (2018), [DOI](https://doi.org/10.1017/pan.2017.44), on preprocessing sensitivity; Rauh (2018), [DOI](https://doi.org/10.1080/19331681.2018.1485608), on sentiment validation in political texts; and Proksch et al. (2019), [DOI](https://doi.org/10.1111/lsq.12218), on political-language measurement across settings.

For another example of dictionary development and testing, see the supplementary lecture case by Boyle et al. (2025), [DOI](https://doi.org/10.1016/j.erss.2025.103955). That study measures policy relevance in research abstracts; our exercise asks whether a sentence discusses energy.

Free reading: [Computational Analysis of Communication](https://cssbook.net/), Chapters 10-11. Technical references: [LIWC Workbench](https://liwc.app/help/workbench), [dictionary import](https://quanteda.io/reference/dictionary.html), [token lookup](https://quanteda.io/reference/tokens_lookup.html).

Selected activities draw on ESS materials by Martijn Schoonvelde (2026), acknowledging Stefan Muller. The parliamentary sample is reused from W2; the short sentences and small dictionaries are course-designed teaching examples.
